---
title: 'Making Data Speak: Principles and Practice of Scientific Visualization'
author: "Ronnie Bailey-Steinitz"
date: "2026-02-06"
output:
  pdf_document: default
  html_document: default
subtitle: "Part 1: Code-Along"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Some useful keyboard shortcuts: 
- New code chunk:  Ctrl+Alt+i // Cmd+Option+i
- Comment or uncomment: Ctrl+Shift+C // Cmd+Shift+C
- Run current selected code line: Ctrl+Enter // Cmd+Return
- Select Word: Ctrl+Shift+Left/Right //	Option+Shift+Left/Right
- Select to Line Start: Alt+Shift+Left // Cmd+Shift+Left
- Select to Line End: Alt+Shift+Right // Cmd+Shift+Right  

## Attach Packages
```{r}

# Core
library(ggplot2)
library(here)

# Data import + wrangling
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(forcats)
library(janitor)

# Plot helpers
library(scales)
library(patchwork)

```

## 0. Read in Data
```{r}

crossings <- readr::read_csv(here("SDSU Data Viz 2026/crossings.csv")) %>% 
  janitor::clean_names()

penguins <- readr::read_csv(here("SDSU Data Viz 2026/penguins.csv")) %>% 
  janitor::clean_names()

```

## 1. Look at data
### 1.1 summary stats
```{r}

dplyr::glimpse(penguins)

# Every ggplot has 3 components: data, aes (aka mapping), geom:
ggplot(penguins, aes(x = bill_length_mm)) +
  geom_histogram()

ggplot(penguins, aes(x = bill_length_mm)) +
  geom_histogram()

```

## 2. Plotting
### 2.1 Scatterplot
```{r}

# BASE
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point()

# Add color
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp)) +
  geom_point()

# # Move legend
# ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp)) +
#   geom_point() +
#   theme(legend.position = "bottom")

# Improve readability - alpha & labels & theme!
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Bill size varies by species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) 

# Custom colors
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Bill size varies by species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen"))

```

> You can find all the keys and hex codes to innate R colors here: https://r-charts.com/colors/  
> ...and more on how to use colors in ggplots here: https://r-graph-gallery.com/ggplot2-color.html  

### 2.2 Adding themes
Themes are pre-defined sets of parameters that give ggplots specific types of appearances
```{r}

ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Bill size varies by species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw()
  # theme_light()
  # theme_dark()
  # theme_minimal()
  # theme_classic() # classic data plot ready for publication
  # theme_void() #excludes anything but the geom itself
  

# you can save the plot and define the dimensions!
ggsave("Bill size by species.png") # if you don't give dimensions, it'll appear like it does in your Plots panel
ggsave("Bill size by species 5x4.png", width = 5, height = 4) # but you can also define dimensions.
# do this to keep multiple plot dimensions consistent in your publication/poster! 

```

> More on themes here: https://ggplot2.tidyverse.org/reference/ggtheme.html

### 2.3 Add trendlines
```{r}

# smooth (curved) lines
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp)) +
  geom_point(alpha = 0.7) +
  geom_smooth() +
  labs(
    title = "Scatterplot + smoothing",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw()

# make it linear!
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm") +
  labs(
    title = "Scatterplot + smoothing",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw()

# Modify the lines
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", linetype = "dashed") +
  labs(
    title = "Scatterplot + smoothing",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw()


# or only have lines:
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp)) +
  geom_smooth(method = "lm", linetype = "dashed") +
  labs(
    title = "Linear Regression",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw()

```

Other linetype options are: twodash; solid; longdash; dotted; dotdash; dashed.

### 2.4 Shapes
```{r}

ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp, shape = spp)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  labs(
    title = "Scatterplot + smoothing",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
  ) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw()


# combine the legends
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp, shape = spp)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  labs(
    title = "Scatterplot + smoothing",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species",
    shape = "Species" # <----- note that we tell ggplot that the legend for the shape has the same name as that for the color, so it auto-combines them
  ) + 
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw()

# ...or...
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = spp, shape = spp)) +
  geom_point(alpha = 0.7, show.legend = FALSE) + # < ---- just tell geom_point NOT to plot legend at all
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  labs(
    title = "Scatterplot + smoothing",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    color = "Species"
    # no "shape" in labels section
  ) + 
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw()

#  Now ggplot knows color and shape belong to the same variable. It automatically merges them into a single legend.

```

### 2.5 Facetting
Also, boxplots and axis text
```{r}

# you can divide by a categorical variable
ggplot(penguins, aes(x = spp, y = body_mass_g, color = spp)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ sex) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw() 

# You can use this to show relationships between three variables: species, sex, body mass

# Boxplots might be a better geom here:
ggplot(penguins, aes(x = spp, y = body_mass_g, color = spp)) +
  geom_jitter(alpha = 0.3, width = 0.2) +
  geom_boxplot(alpha = 0.5) +
  facet_wrap(~ sex) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```

To use `vjust` and `hjust` for axis text in ggplot2, modify the `theme()` function with `axis.text.x` or `axis.text.y` and the `element_text()` function. The values for `vjust` (vertical justification) and hjust (horizontal justification) range from 0 to 1. 

Next, we'll look at how to manage your viewer's attention, and in the process, we'll see new types of charts.

## 3. Storytelling: Managing the viewer's attention
### 3.1 Using colors
```{r}

# let's look at some other aspects of the data
penguins %>% dplyr::count(clutch_completion, spp)

# basic stacked bar chart
ggplot(penguins, aes(x = spp, fill = clutch_completion)) +
  geom_bar() +
  
  # scale_fill_manual(values = c("green4", "firebrick3")) + 
      # but notice the colors, what are they telling you???
  
  # scale_fill_manual(values = c("firebrick", "green4")) + 
      # this makes more sense-- play on peoples' familiarity with color-coding!!! 
  
  labs(x = "Species", y = "N") +
  theme_bw() 


# you can change the focus of the chart (swap color vs. x-axis)
ggplot(penguins, aes(x = clutch_completion, fill = spp)) +
  geom_bar() +
  #scale_fill_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  labs(x = "Clutch Completion", y = "N") +
  theme_bw() 

# --> we simply switched which variable was on the x-axis, and while is the color fill, but it makes a difference in how you tell the story!



# To draw the viewer's attention, you can also highlight one particular aspect using colors
ggplot(penguins, aes(x = clutch_completion, fill = spp)) +
  geom_bar() +
  scale_fill_manual(values = c("darkorange3", "grey80", "grey90")) +
  labs(x = "Clutch Completion", y = "N") +
  theme_bw() 

#########
# ADD N = numbers to bars???
#########

# you can also go back to you summary stats (the histogram):
ggplot(penguins, aes(x = bill_length_mm, fill = spp)) +
  geom_histogram() +
  labs(x = "Bill Length (mm)", y = "N", fill = "Species") +
  theme_bw()

ggplot(penguins, aes(x = bill_length_mm, fill = spp)) +
  geom_histogram() +
  scale_fill_manual(values = c("darkorange3", "grey80", "grey90")) +
  labs(x = "Bill Length (mm)", y = "N", fill = "Species") +
  theme_bw()

# ======================================================

plot1 <- ggplot(penguins, aes(x = clutch_completion, fill = spp)) +
  geom_bar() +
  scale_fill_manual(values = c("darkorange3", "grey80", "grey90")) +
  labs(x = "Clutch Completion", y = "N", fill = "Species") +
  theme_bw() +
  theme(panel.grid.major.x = element_blank())

plot2 <- ggplot(penguins, aes(x = bill_length_mm, fill = spp)) +
  geom_histogram() +
  scale_fill_manual(values = c("darkorange3", "grey80", "grey90")) +
  labs(x = "Bill Length (mm)", y = "N", fill = "Species") +
  theme_bw()

# this is part of the package {patchwork}
plot1 + plot2 + plot_layout(axes = "collect", guides = "collect")

```

### 3.2 Using visual cues
```{r}
# let's look at bill length by egg hatching date 'date_e'
ggplot(penguins, aes(x = date_e, y = bill_length_mm)) +
  geom_point()

# hmm... looks like there was one season each year from 2007-2009...
# so, does it make sense to plot the x-axis as full dates? with months and days?

# seems like a waste of space.

# instead, we can focus on what we want to look at-- egg-laying season, which seems to be days within the month of November in these years
ggplot(penguins, aes(x = lubridate::day(date_e), y = bill_length_mm)) +
  geom_point()

# good, but not there yet...

# let's try to see what's happening by species:
# add 'color = spp' to aes() in ggplot
ggplot(penguins, aes(x = lubridate::day(date_e), y = bill_length_mm, color = spp)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  labs(x = "Hatching Day, Month of November", y = "Bill Length (mm)") +
  # scale_x_continuous(breaks = c(1:30)) +
  theme_bw() 



# let's say something happened on a particular date that you want to highlight:
# for example, if a few days had rain:
ggplot(penguins, aes(x = lubridate::day(date_e), y = bill_length_mm, color = spp)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  annotate("rect", 
           xmin = 20, # from the 20th
           xmax = 22, # to the 22nd
           ymin = Inf, # set ymin to +infinity
           ymax = -Inf, # set ymax to -infinity 
           fill = "red", 
           # fill = "skyblue",  # maybe use a color that makes sense for rain instead of red
           alpha = 0.3) +
  geom_point(data = penguins %>% filter(individual_id == "N16A1"), color = "red", size = 5) +
  # annotate("text", label = "<-- Monsoon", x = 25, y = 58, color = "blue") + # include text annotation if you want
  labs(x = "Hatching Day, Month of November", y = "Bill Length (mm)") +
  theme_bw() 



# or if you want to highlight a threshold, use a single line (instead of a rectangle):
ggplot(penguins, aes(x = lubridate::day(date_e), y = bill_length_mm, color = spp)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("purple4", "orchid3", "lightseagreen")) +
  geom_hline(yintercept = 50, color = "green4", linetype = "dashed", linewidth = 1.5, alpha = 0.7) +
  # geom_vline(xintercept = 15, color = "red", linetype = "solid", linewidth = 1.5, alpha = 0.7) +
  labs(x = "Hatching Day, Month of November", y = "Bill Length (mm)") +
  theme_bw() 

```


## 4. Primate Data
### 4.1 Explore
```{r}

# let's have a look at the data:
glimpse(crossings)

# Ok, no numeric variables. So, I'd want to compile them somehow.
# Let's group by a/s class, and count the number of crossings per a/s class
# and then convert that to a proportion
# Completion rate by sex
cross1 <- crossings %>%
  # dplyr::filter(!is.na(age_sex_class)) %>%
  group_by(age_sex_class) %>%
  summarise(
    n_crossings = n(),
    prop_complete = mean(cross_complete == "C", na.rm = TRUE)
  )

# This will give:
# Count per age class
# Proportion complete (numeric 0–1)

# plot:
# Bar chart of prop_complete
# Add n_crossings as text labels
ggplot(cross1, aes(x = age_sex_class, y = prop_complete)) +
  geom_col() +
  geom_text(aes(label = round(prop_complete, 3)), vjust = -0.6, size = 3.5) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Crossing completion rate by age/sex class",
    x = "Age/Sex/Class",
    y = "Proportion complete"
  ) +
  theme_bw()

# add: %>% filter(!is.na(age_sex_class)) # to wrangling

# ok, let's look at how 
# Gait distribution by group
cross2 <- crossings %>%
  group_by(group, cross_gait) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(group) %>%
  mutate(prop = n / sum(n))

ggplot(cross2, aes(x = group, y = prop, fill = cross_gait)) +
  geom_col(width = 0.8) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Crossing gait distribution by group",
    x = "Group",
    y = "Proportion of crossings",
    fill = "Cross gait"
  ) +
  theme_bw()

```


## 5. Bonus content
```{r}

# heat map; 
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_density2d_filled(color = "black", linewidth = 0.2) +
  facet_wrap(~ spp) +
  labs(
    title = "Overplotting solution: Heat Mat",
    subtitle = "Colors depict the number of points that fall in that region",
    x = "Bill length (mm)",
    y = "Bill depth (mm)",
    fill = "Count"
  ) +
  theme_bw()

```


